---
title: "Exercise Quality Prediction from Accelerometer Data"
author: "catharpin"
date: "November 8, 2015"
output: html_document
---

# Executive Summary

Using accelerometer and position sensor data collected from test subjects performing physical exercises in a controlled environment, 
a machine learning algorithm is developed to predict whether
these exercises are being done correctly (and, if incorrect, if the incorrect performance can be specifically categorized.)

Using a boosting with trees algorithm, using K-fold cross validation (n=10), a model is developed that with an error rate of 3.93%, with a 95% confidence interval of (3.34%,4.58%).  This error estimation was determined by withholding 20% of the training data by random sampling.

# Libraries Used

```{r libraries, results='hide'}
library(doMC)
library(caret)
registerDoMC(cores=8)
set.seed(10191) 
```

# Retrieving Data

The data used is from the "Weight Lifting Exercises Dataset", retrieved 08 NOV 2015 from http://groupware.les.inf.puc-rio.br/har#dataset [1].

```{r data-retrieve-load,echo=FALSE}
if (!file.exists("./data")){dir.create("./data")}

if (!file.exists("./data/train.csv")){
  message("Retrieving training data set.")
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",destfile = "./data/train.csv")
} else {
  message("Training data found.")
}

if (!file.exists("./data/test.csv")){
  message("Retrieving testing data set.")
  download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile="./data/test.csv")
} else {
  message("Testing data found.")
}

orig_train <- read.csv("./data/train.csv")
orig_test  <- read.csv("./data/test.csv")
```

#Data Cleansing

Many of the variables are entirely missing from the test set and as a result have no predictive value.  These are removed from both training and test sets to increase tidiness and reduce effort of the fitting algorithm; we also remove time-of-day and data-collection artifacts.

Initial fitting of the data showed the 'num_window' variable (column 7 in the data) as highly influential.  
While it improved the accuracy of the model, in the author's opinion it is not a valid predictor in the spirit of the effort, 
as it is an artifact of how the data was collected, and was removed.



```{r cleansing-1}
# Remove near-zero variables (from the testing set) from data frames.
nzv_test <- nearZeroVar(orig_test)
test <- orig_test[,-nzv_test]
train <- orig_train[,-nzv_test]

# Remove data-collection artifacts from data frames.
test <- test[,c(-1,-3:-6)]
train <- train[,c(-1,-3:-6)]
```



# Data Segmentation

To provide an independent evaluation of the out-of-sample error rate, the training sample is further subdivided into training and 'self-test'
sets.

```{r segmentation-1}
inTrain <- createDataPartition(y = train$classe, p=0.8, list=FALSE)
train2<-train[inTrain,]
self_test<-train[-inTrain,]
```

# Model Fit

Some of the data could not be converted to numeric values, causing the fit function to emit a warning; 
I have chosen to leave those as NA rather than to impute them, as it wasn't 
necessary to obtain a high-accuracy fit.  

```{r fit-1, cache=TRUE, results='hide'}
tc <- trainControl(method="cv", number=10)
fit1<-train(classe ~ ., data=train2, preProcess=c("center","scale"), method="gbm", trControl = tc, metric="Accuracy", verbose="FALSE")
```

# Result

```{r fit-confusion}
confusionMatrix(fit1)
```

# Out of Sample Error Estimation

To estimate the error, we perform a fit on the 'self-test' set and determine the error rate.  See Accuracy, below, for the estimated
accuracy and confidence interval.

```{r error-estimate, results='show'}
error_prediction <- predict(fit1, self_test)
cm_outsample<-confusionMatrix(error_prediction, self_test$classe)
cm_outsample
```

```{r graphic-attempt, echo=FALSE}
rgb.palette <- colorRampPalette(c("white","yellow","red","black"), space="rgb")
mypanel <- function(x,y,z,...) { panel.levelplot(x,y,z,...); panel.text(x,y,round(z,3))}
levelplot(cm_outsample$table/rowSums(cm_outsample$table),panel=mypanel,col.regions=rgb.palette(120),cuts=100,at=seq(0,1.2,.01),xla="Predicted Class",yla="Actual Class",main="Normalized Confusion Matrix")

```

# Prediction

The model is finally applied to the test dataset for evaluation.

```{r prediction}
pred <- predict(fit1, test)
pred
```

```{r write-results,eval=FALSE, echo=FALSE,results='hide'}

# This function is (as recommended by jtleek) to simplify generation of files for submission for the assignment.
# It's not necessary for the analysis so it's commented out here.

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("jh_problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

# pml_write_files(as.character(pred))

```



# Bibliography

[1] Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. _Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13)_ . Stuttgart, Germany: ACM SIGCHI, 2013. 

